\texttt{dfoliatR}: An R package for detection and analysis of insect defoliation signals in tree rings

### 2020-09-02 README.Rmd
update readme
updates for proof and publication

# `dfoliatR`: An R package for detection and analysis of insect defoliation signals in tree rings

@@

A repository for the initial descriptive paper of the [`dfoliatR`](https://chguiterman.github.io/dfoliatR/) package in R. The paper is currently being prepared for submission to *Dendrochronologia*.

```{r, echo=FALSE}
knitr::include_graphics("dfoliatR_ms.pdf")
```

Please contact Chris Guiterman with questions or comments.

>>>>

### `dfoliatR`: An R package for detection and analysis of insect defoliation signals in tree rings

@@

This repository includes the code and text to generate a preprint version of the published paper:

>Guiterman, CH, AM Lynch, and JN Axelson (2020) [`dfoliatR`: An R package for detection and analysis of insect defoliation signals in tree rings.](https://authors.elsevier.com/sd/article/S1125786520300898) _Dendrochronologia_. DOI: 10.1016/j.dendro.2020.125750.

The paper provides an overview of the `dfoliatR` R package. More information on using the software is availabe from [https://chguiterman.github.io/dfoliatR/](https://chguiterman.github.io/dfoliatR/).

Please contact @chguiterman with questions or comments.

### 2020-09-02 dfoliatR_ms.Rmd
updates for proof and publication

```{r setup, include=FALSE}
library(dfoliatR)
library(knitr)
library(magrittr)
library(here)
# library(kableExtra)
# options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
# # opts_chunk$set(echo = TRUE)
# # knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', `\\usepackage[table]{xcolor}', x, fixed = TRUE)})`)})
# knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
```
> This article is published: Guiterman, CH, AM Lynch, and JN Axelson (2020) `dfoliatR`: An R package for detection and analysis of insect defoliation signals in tree rings. _Dendrochronologia_. DOI: 10.1016/j.dendro.2020.125750.
> Available Open Access from [https://authors.elsevier.com/sd/article/S1125786520300898](https://authors.elsevier.com/sd/article/S1125786520300898)

@@

```{r echo=TRUE, eval=FALSE}
dmj_defol_stats <- defol_stats(dmj_defol)
View(dmj_defol_stats)
```

@@

```{r echo=TRUE, eval=FALSE}
plot_defol(dmj_defol)
## This creates a ggplot object. Additions can 
## be made to adjust plotting themes and
## aesthetics, like color.   
## For more on the features of ggplot see
## https://ggplot2.tidyverse.org/
## Adding 'scale_color_manual()' will
## change the colors that denote 
## the average NGSI value for each defoliation
## event.
library(ggplot2)
plot_defol(dmj_defol) +
  scale_color_manual(values = c("red", "orange", "purple"))
```

@@

Input parameters to `outbreak()` include "filter_min_series" to control the chronology cut-off points with regard to sample depth, "filter_min_defol" and "filter_perc" to control the minimum number and percent of trees recording a defoliation event in a given year. `outbreak()` produces a new data frame with eight variables: "year", "num_defol", "percent_defol", "num_max_defol", "mean_gsi", "mean_ngsi", and "outbreak_status." All of these variables are populated regardless of an inferred outbreak event, providing a continuous outbreak reconstruction. The "num_max_defol" variable counts the number of trees recording their maximum defoliation in a given year. The "mean_gsi" and "mean_ngsi" variables provide averages of these indices across all available trees. Finally, the "outbreak_status" column shows whether an outbreak event is inferred ("outbreak") or not ("not_obr").

@@

```{r echo=TRUE, eval=FALSE}
plot_outbreak(dmj_obr, disp_index = "GSI")
```

@@

```{r echo=TRUE, eval=FALSE}
dmj_obr_stats <- outbreak_stats(dmj_obr)
View(dmj_obr_stats)
```

@@

```{r echo=TRUE, eval=FALSE}
## calculate mean return interval for outbreak start years
intervals <- diff(dmj_obr_stats$start)
mean(intervals) ## 51.2 year mean return interval
```

>>>>

```{r setup, include=FALSE}
library(dfoliatR)
library(knitr)
library(magrittr)
library(here)
# library(kableExtra)
# options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
# # opts_chunk$set(echo = TRUE)
# # knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', `\\usepackage[table]{xcolor}', x, fixed = TRUE)})`)})
# knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
```
> This article is published as: Guiterman, CH, AM Lynch, and JN Axelson (2020) `dfoliatR`: An R package for detection and analysis of insect defoliation signals in tree rings. _Dendrochronologia_. DOI: 10.1016/j.dendro.2020.125750. Available for free [(link)](https://authors.elsevier.com/sd/article/S1125786520300898)

@@

```{r echo=TRUE, eval=FALSE}
dmj_defol_stats <- defol_stats(dmj_defol)
View(dmj_defol_stats) # Table 1
```

@@

```{r echo=TRUE, eval=FALSE}
plot_defol(dmj_defol) # Fig. 1
## This creates a ggplot object. Additions can 
## be made to adjust plotting themes and
## aesthetics, like color.   
## For more on the features of ggplot see
## https://ggplot2.tidyverse.org/
## Adding 'scale_color_manual()' will
## change the colors that denote 
## the average NGSI value for each defoliation
## event.
library(ggplot2)
plot_defol(dmj_defol) +
  scale_color_manual(values = c("red", "orange", 
                                "purple"))
```

@@

Input parameters to `outbreak()` include "filter_min_series" to control the chronology cut-off points with regard to sample depth, "filter_min_defol" and "filter_perc" to control the minimum number and percent of trees recording a defoliation event in a given year. `outbreak()` produces a new data frame with eight variables: "year", "num_defol", "percent_defol", "num_max_defol", "mean_gsi", "mean_ngsi", and "outbreak_status." All of these variables are populated regardless of an inferred outbreak event, providing a continuous outbreak reconstruction. The "num_max_defol" variable counts the number of trees recording their maximum defoliation in a given year. The "mean_gsi" and "mean_ngsi" variables provide averages of these indices across all available trees. Finally, the "outbreak_status" column shows if an outbreak event is inferred ("outbreak") or not ("not_obr"), and whether it represents an ongoing series-end event ("se_outbreak").

@@

```{r echo=TRUE, eval=FALSE}
plot_outbreak(dmj_obr, disp_index = "GSI") # Fig. 2
```

@@

```{r echo=TRUE, eval=FALSE}
dmj_obr_stats <- outbreak_stats(dmj_obr)
View(dmj_obr_stats) # Table 2
```

@@

```{r echo=TRUE, eval=FALSE}
## calculate mean return interval for outbreak 
## start years
intervals <- diff(dmj_obr_stats$start)
mean(intervals) ## 51.2 year mean return interval
```

### 2020-09-02 dfoliatR_ms.tex
updates for proof and publication

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dmj_defol_stats <-}\StringTok{ }\KeywordTok{defol_stats}\NormalTok{(dmj_defol)}
\KeywordTok{View}\NormalTok{(dmj_defol_stats)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_defol}\NormalTok{(dmj_defol)}
\CommentTok{## This creates a ggplot object. Additions can }
\CommentTok{## be made to adjust plotting themes and}
\CommentTok{## aesthetics, like color.   }
\CommentTok{## For more on the features of ggplot see}
\CommentTok{## https://ggplot2.tidyverse.org/}
\CommentTok{## Adding 'scale_color_manual()' will}
\CommentTok{## change the colors that denote }
\CommentTok{## the average NGSI value for each defoliation}
\CommentTok{## event.}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{plot_defol}\NormalTok{(dmj_defol) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"red"}\NormalTok{, }\StringTok{"orange"}\NormalTok{, }\StringTok{"purple"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

Input parameters to \texttt{outbreak()} include ``filter\_min\_series'' to control the chronology cut-off points with regard to sample depth, ``filter\_min\_defol'' and ``filter\_perc'' to control the minimum number and percent of trees recording a defoliation event in a given year. \texttt{outbreak()} produces a new data frame with eight variables: ``year'', ``num\_defol'', ``percent\_defol'', ``num\_max\_defol'', ``mean\_gsi'', ``mean\_ngsi'', and ``outbreak\_status.'' All of these variables are populated regardless of an inferred outbreak event, providing a continuous outbreak reconstruction. The ``num\_max\_defol'' variable counts the number of trees recording their maximum defoliation in a given year. The ``mean\_gsi'' and ``mean\_ngsi'' variables provide averages of these indices across all available trees. Finally, the ``outbreak\_status'' column shows whether an outbreak event is inferred (``outbreak'') or not (``not\_obr'').

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_outbreak}\NormalTok{(dmj_obr, }\DataTypeTok{disp_index =} \StringTok{"GSI"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dmj_obr_stats <-}\StringTok{ }\KeywordTok{outbreak_stats}\NormalTok{(dmj_obr)}
\KeywordTok{View}\NormalTok{(dmj_obr_stats)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## calculate mean return interval for outbreak start years}
\NormalTok{intervals <-}\StringTok{ }\KeywordTok{diff}\NormalTok{(dmj_obr_stats}\OperatorTok{$}\NormalTok{start)}
\KeywordTok{mean}\NormalTok{(intervals) }\CommentTok{## 51.2 year mean return interval}
\end{Highlighting}
\end{Shaded}

>>>>

\begin{quote}
This article is published as: Guiterman, CH, AM Lynch, and JN Axelson (2020) \texttt{dfoliatR}: An R package for detection and analysis of insect defoliation signals in tree rings. \emph{Dendrochronologia}. DOI: 10.1016/j.dendro.2020.125750. Available for free \href{https://authors.elsevier.com/sd/article/S1125786520300898}{(link)}
\end{quote}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dmj_defol_stats <-}\StringTok{ }\KeywordTok{defol_stats}\NormalTok{(dmj_defol)}
\KeywordTok{View}\NormalTok{(dmj_defol_stats) }\CommentTok{# Table 1}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_defol}\NormalTok{(dmj_defol) }\CommentTok{# Fig. 1}
\CommentTok{## This creates a ggplot object. Additions can }
\CommentTok{## be made to adjust plotting themes and}
\CommentTok{## aesthetics, like color.   }
\CommentTok{## For more on the features of ggplot see}
\CommentTok{## https://ggplot2.tidyverse.org/}
\CommentTok{## Adding 'scale_color_manual()' will}
\CommentTok{## change the colors that denote }
\CommentTok{## the average NGSI value for each defoliation}
\CommentTok{## event.}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{plot_defol}\NormalTok{(dmj_defol) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"red"}\NormalTok{, }\StringTok{"orange"}\NormalTok{, }
                                \StringTok{"purple"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

Input parameters to \texttt{outbreak()} include ``filter\_min\_series'' to control the chronology cut-off points with regard to sample depth, ``filter\_min\_defol'' and ``filter\_perc'' to control the minimum number and percent of trees recording a defoliation event in a given year. \texttt{outbreak()} produces a new data frame with eight variables: ``year'', ``num\_defol'', ``percent\_defol'', ``num\_max\_defol'', ``mean\_gsi'', ``mean\_ngsi'', and ``outbreak\_status.'' All of these variables are populated regardless of an inferred outbreak event, providing a continuous outbreak reconstruction. The ``num\_max\_defol'' variable counts the number of trees recording their maximum defoliation in a given year. The ``mean\_gsi'' and ``mean\_ngsi'' variables provide averages of these indices across all available trees. Finally, the ``outbreak\_status'' column shows if an outbreak event is inferred (``outbreak'') or not (``not\_obr''), and whether it represents an ongoing series-end event (``se\_outbreak'').

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_outbreak}\NormalTok{(dmj_obr, }\DataTypeTok{disp_index =} \StringTok{"GSI"}\NormalTok{) }\CommentTok{# Fig. 2}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dmj_obr_stats <-}\StringTok{ }\KeywordTok{outbreak_stats}\NormalTok{(dmj_obr)}
\KeywordTok{View}\NormalTok{(dmj_obr_stats) }\CommentTok{# Table 2}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## calculate mean return interval for outbreak }
\CommentTok{## start years}
\NormalTok{intervals <-}\StringTok{ }\KeywordTok{diff}\NormalTok{(dmj_obr_stats}\OperatorTok{$}\NormalTok{start)}
\KeywordTok{mean}\NormalTok{(intervals) }\CommentTok{## 51.2 year mean return interval}
\end{Highlighting}
\end{Shaded}
